\propertitle

### 2019-12-20 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-20T11:10:29+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As a first step we need to translate vague terms like
\enquote{cooperativity}, \enquote{interaction}, and similar into a more
precise, operational \citep[\chap~I]{bridgman1927_r1958} notion. Here we
propose the notion of \emph{informational sufficiency} (also called
\enquote{predictive sufficiency})
\citep{kullbacketal1951}[\sect~4.5]{bernardoetal1994}[\chap~8 and
\sect~14.2]{jaynes1994_r2003}, because
\begin{enumerate*}[label=(\roman*)]
\item it seems closely related to those terms as used in the
  literature; \item it is mathematically connected with maximum-entropy
  distributions; \item it reveals important consequences of sampling.
\end{enumerate*}
Its meaning and the three points above will now be illustrated; please see
the end of this section for references to in-depth discussions and proofs
of technical results.

@@

But such a translation does more: through a mathematical theorem, it leads
to the automatic appearance of maximum-entropy distributions: \emph{if a
  probability distribution for a repetitive phenomenon has some sufficient
  statistics, then it is a mixture of distributions of maximum-entropy form
  having those statistics as constraints, and vice versa}. See the end of
this section for detailed references about this theorem.
% Thus, maximum-entropy distributions automatically appear (as
% approximations) from the probability-calculus when we test hypotheses
% about informational sufficiency.
Its most important consequence is that, by Bayes's formula, the probability
that some statistics are informationally sufficient can be approximated by
finding the maximum-entropy distribution constrained by those statistics
and then calculating the probability for the whole data under that
distribution. We shall shortly illustrate this kind of calculation for the
first data set.

@@

We close this section with some references. The notion of informational or
predictive sufficiency, probably first presented by Kullback \amp\ Leibler
\parencite*{kullbacketal1951}, is discussed in Bernardo \etal\
\parencite*[\sect~4.5]{bernardoetal1994}, who explain its connections with
Fisher's\parencite*{fisher1922} parametric sufficiency. Jaynes
\parencite*[\chap~8 and \sect~14.2] shows the connections with information
theory. The mathematical relation between sufficiency and the exponential
family comes from the Pitman-Koopman
theorem \citep{koopman1936,pitman1936,darmois1935} for continuous variables.

\mynote{Short reasoned bibliography about sufficient statistics and its
  relation to maximum-entropy?} blah blah \footnote{For more technical results and
  connections with the notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,lauritzen1982_r1988,lauritzen2007,cifarellietal1980,cifarellietal1981,diaconisetal1981,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,ayetal2015}.}
This mathematical result comes from the Pitman-Koopman
theorem\citep{cifarellietal1982,kullbacketal1951}[the
notion goes back to][]{fisher1922} \citep{jaynes1982b}[\sect~4.5.4]{bernardoetal1994}
\citep{koopman1936,pitman1936,darmois1935}[for later analyses and the
discrete case
see][]{hipp1974,andersen1970,denny1967,denny1972,fraser1963,barankinetal1963,barndorffnielsen1978_r2014}

@@

The crucial point is our belief distribution for the activity of the sample
\emph{conditional} on the activity of the larger population. It is given by
the hypergeometric distribution~\eqref{eq:conditional_hypergeometric},
reprinted here:
\begin{equation}
  \labelbis{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
    \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}.
\end{equation}
This distribution leads to the equality of factorial
moments~\eqref{eq:equal_m_factorialmoments}, on which our model rests. This
conditional probability, characteristic of \enquote{drawing without
  replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968},
ensues when our degree of belief that the $i$th unit (ball or neuron) will
be drawn has the same value for all $i \in \set{1,\dotsc,N}$. Thus,
consider the pool of all neurons which we believe -- with equal degree for
each neuron -- could have been recorded. $N$ is the size of that pool. $N$
therefore depends on factors such as: the shape, dimensions, and technical
specifications of the recording probe; the inaccuracy in the insertion of
the probe, leading for example to slightly different insertion points or
angles; the density of neurons around the probe. But it also depends on the
homogeneity of the brain region where the recording was made: if we believe
that it doesn't matter whether the probe had been inserted in some other
point of the same brain region, then
formula~\eqref{eq:conditional_hypergeometric} is appropriate. $N$ can
therefore only be assessed case by case.

Whether such equal beliefs are justified, or for which sampling procedures
they can be justified, is the fundamental, deep question of sampling
theory, which we cannot discuss here \citep[see \eg\ the discussions,
reviews, and references
in][]{ericson1969,smith1976,kruskaletal1979c,kruskaletal1980}. It is of
course possible to probabilistically assess, case by case, whether this
equality assumption is appropriate. But we must also be aware that such
assessment in turn rests on analogous assumptions at a higher level, and so
on. The probability-calculus, just like the truth-calculus, cannot yield
conclusions without premisses \citep[p.~182]{johnson1924}.

@@

The first and most important assumption, typical of this kind of
maximum-entropy application in neuroscience, can be expressed in two
equivalent ways:
\begin{enumerate*}[(\roman*)]
  % \item the order of the time bins has no relevance;
\item the frequencies of the activity during a recording are
  informationally sufficient for inferences about unrecorded times; \item
  our degree of belief about the sequence of activities is invariant under
  permutations of the time bins.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
is to relate the probability for the frequency distribution $\yF$ during
the recording to that of the \emph{long-run} frequency distribution for the
activities of the larger population. Note that the maximum-entropy method
in statistical physics is meant to be used not with this assumption, but
with averages across \emph{repetitions} of the same experiment, because its
constraints represent macroscopically \emph{reproducible} quantities and
kinematics, even in non-equilibrium
\citep{jaynes1957,guntonetal1983,tikochinskyetal1984,grandy1988,berryetal1988,deroecketal2005_r2006,touchette2009}.

@@

First of all the sampling relation. We have a population of $N$ units --
neurons -- $\yA$ of which have some specific property -- being active in a
specific time bin -- and we sample in an unknown way $n$ of the $N$ units.
The probability that $\ya$ of the $n$ sampled neurons are active is then
given by the hypergeometric distribution
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
  \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv  \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}
\end{equation}
typical of \enquote{drawing without replacement}
\citep[\chap~3]{jaynes1994_r2003}[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}[\sects~2.1,
3.2]{jeffreys1939_r1983}. In the following we leave $n$, $N$ implicit in
the conditional. If we are uncertain about the number $\yA$, with belief
$P(\yA)$, then by the theorem of total probability our belief about $\ya$
is
\begin{equation}
  \label{eq:belief_sample_unknown_A}
  p(\ya)= \sum_{\yA} \yG_{\ya\yA} \;P(\yA).
\end{equation}

From the definition of normalized factorial
moment~\eqref{eq:def_factorial_moment_a}, the expression for the
hypergeometric distribution~\eqref{eq:conditional_hypergeometric}, and the
relation~\eqref{eq:belief_sample_unknown_A} between our beliefs about $\ya$
and $\yA$, and using some combinatorial juggling
\citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[appendix~A]{portamanaetal2015}{potts1953}, one
can prove the equality~\eqref{eq:equal_m_factorialmoments} between the
factorial moments of sample and larger population.

@@

For the construction of a maximum-entropy distribution from generic
expectation constraints see Jaynes
\citey{jaynes1963}[\chap~11]{jaynes1994_r2003}. More precisely we use the
minimum relative-entropy method
\citep{hobsonetal1973,csiszar1985}[\sect~5.2.2]{sivia1996_r2006} with
respect to a reference distribution. For the solution of the extremization
problem using Lagrangians and Lagrange multipliers see Mead \etal\
\citey{meadetal1984} and the extensive texts by Fang \etal\
\citey{fangetal1997} and Boyd \etal\ \citey{boydetal2004_r2009}. For a
geometric understanding of the extremization and of the relation between
expectations and multipliers see Porta Mana \citey{portamana2017b}.

@@

The first assumption, of infinite exchangeability
\citep[\chap~4]{bernardoetal1994_r2000}{dawid2013} of the probability
distribution $\pf[(\yA_{1},\dotsc, \yA_{T}) \| \yI]$ for the sequence of
activities, implies the following mixture form by de~Finetti's theorem
\citep{definetti1930,hewittetal1955}:
\begin{equation}
  \label{eq:mixture_from_definetti}
%  \exp[-T\;\sh(\yF;\ynu)] \quad
%  \pxxx[(\yA_{1},\dotsc,\yA_{T})] \propto
  \pf[(\yA_{1},\dotsc, \yA_{T}) \| \yI]
  = \int\!\!\!\di\ynu\;\pf(\ynu \|\yI)\,\prod_{\yA=0}^{N}{\nu_{\yA}}^{T\,\yFF_{\yA}},
\end{equation}
where $\ynu\defd(\nu_{\yA})$ is the long-run frequency distribution of
activities in an imaginary continuation or repetition of the recording, and
$\pf(\ynu \|\yI)\;\di\ynu$ our belief distribution about it. The
integration is over an $N$-dimensional simplex.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-20T20:13:38+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

As a first step we need to translate vague terms like
\enquote{cooperativity}, \enquote{interaction}, and similar into a more
precise, operational \citep[\cf][\chap~I]{bridgman1927_r1958} notion. Here
we propose the notion of \emph{informational sufficiency}, because
\begin{enumerate*}[label=(\roman*)]
\item it seems closely related to those terms as used in the
  literature; \item it is mathematically connected with maximum-entropy
  distributions; \item it reveals important consequences of sampling.
\end{enumerate*}
Its meaning and the three points above will now be illustrated. For an
extensive discussion of this notion, probably first suggested by Kullback
\amp\ Leibler \parencite*{kullbacketal1951}, see Bernardo \etal\
\parencite*[\sect~4.5]{bernardoetal1994} and Jaynes \parencite*[\chap~8 and
\sect~14.2]{jaynes1994_r2003}.

@@

But such a translation does more: through a mathematical theorem \citep[and
references therein]{andersen1970}[this theorem was first proved for
continuous probability distributions
by][]{koopman1936,pitman1936,darmois1935}[see][\sect~4.5.3 for further
references]{bernardoetal1994}, it leads to the automatic appearance of
maximum-entropy distributions: \emph{if a probability distribution for a
  repetitive phenomenon has some sufficient statistics, then it is a
  mixture of distributions of maximum-entropy form having those statistics
  as constraints, and vice versa}. 
% Thus, maximum-entropy distributions automatically appear (as
% approximations) from the probability-calculus when we test hypotheses
% about informational sufficiency.
Its most important consequence is that, by Bayes's formula, the probability
that some statistics are informationally sufficient can be approximated by
finding the maximum-entropy distribution constrained by those statistics
and then calculating the probability for the whole data under that
distribution. We shall shortly illustrate this kind of calculation for the
first data set.

@@

% We close this section with some references. The notion of informational or
% predictive sufficiency, probably first presented by Kullback \amp\ Leibler
% \parencite*{kullbacketal1951}, is discussed in Bernardo \etal\
% \parencite*[\sect~4.5]{bernardoetal1994}, who explain its connections with
% Fisher's\parencite*{fisher1922} parametric sufficiency. Jaynes
% \parencite*[\chap~8 and \sect~14.2] shows the connections with information
% theory. The mathematical relation between sufficiency and the exponential
% family comes from the Pitman-Koopman
% theorem \citep{koopman1936,pitman1936,darmois1935} for continuous variables.

% \mynote{Short reasoned bibliography about sufficient statistics and its
%   relation to maximum-entropy?} blah blah \footnote{For more technical results and
%   connections with the notion of symmetry see \eg\
%   \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,lauritzen1982_r1988,lauritzen2007,cifarellietal1980,cifarellietal1981,diaconisetal1981,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,ayetal2015}.}
% This mathematical result comes from the Pitman-Koopman
% theorem\citep{kullbacketal1951}[the
% notion goes back to][]{fisher1922} \citep{jaynes1982b}[\sect~4.5.4]{bernardoetal1994}
% \citep{koopman1936,pitman1936,darmois1935}[for later analyses and the
% discrete case
% see][]{andersen1970,%discrete
% denny1967,denny1972}

@@

The crucial point is our belief distribution for the activity of the sample
\emph{conditional} on the activity of the larger population. It is given by
the hypergeometric distribution~\eqref{eq:conditional_hypergeometric},
reprinted here:
\begin{equation}
  \labelbis{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
    \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}.
\end{equation}
This distribution leads to the equality of factorial
moments~\eqref{eq:equal_m_factorialmoments}, on which our model rests. This
conditional probability, characteristic of \enquote{drawing without
  replacement}
\citep[\eg][\chap~3]{jaynes1994_r2003}, % [\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}
ensues when our degree of belief that the $i$th unit (ball or neuron) will
be drawn has the same value for all $i \in \set{1,\dotsc,N}$. Thus,
consider the pool of all neurons which we believe -- with equal degree for
each neuron -- could have been recorded. $N$ is the size of that pool. $N$
therefore depends on factors such as: the shape, dimensions, and technical
specifications of the recording probe; the inaccuracy in the insertion of
the probe, leading for example to slightly different insertion points or
angles; the density of neurons around the probe. But it also depends on the
homogeneity of the brain region where the recording was made: if we believe
that it doesn't matter whether the probe had been inserted in some other
point of the same brain region, then
formula~\eqref{eq:conditional_hypergeometric} is appropriate. $N$ can
therefore only be assessed case by case.

Whether such equal beliefs are justified, or for which sampling procedures
they can be justified, is the fundamental, deep question of sampling
theory, which we cannot discuss here \citep[see \eg\ the discussions and
references
in][]{ericson1969,smith1976}. % {kruskaletal1979c,kruskaletal1980}
It is of course possible to probabilistically assess, case by case, whether
this equality assumption is appropriate. But we must also be aware that
such assessment in turn rests on analogous assumptions at a higher level,
and so on. The probability-calculus, just like the truth-calculus, cannot
yield conclusions without premisses \citep{hailperin2011}[p.~182]{johnson1924}.

@@

The first and most important assumption, typical of this kind of
maximum-entropy application in neuroscience, can be expressed in two
equivalent ways:
\begin{enumerate*}[(\roman*)]
  % \item the order of the time bins has no relevance;
\item the frequencies of the activity during a recording are
  informationally sufficient for inferences about unrecorded times; \item
  our degree of belief about the sequence of activities is invariant under
  permutations of the time bins.
\end{enumerate*}
Equating time averages and expectations, \eqn~\eqref{eq:final_constraint},
would not make sense without this assumption. Its mathematical consequence
is to relate the probability for the frequency distribution $\yF$ during
the recording to that of the \emph{long-run} frequency distribution for the
activities of the larger population. Note that the maximum-entropy method
in statistical physics is meant to be used not with this assumption, but
with averages across \emph{repetitions} of the same experiment, because its
constraints represent macroscopically \emph{reproducible} quantities and
kinematics, even in non-equilibrium
\citep{jaynes1985b_r1996,%jaynes1957,%tikochinskyetal1984,
grandy1988,%berryetal1988,
deroecketal2005_r2006%,touchette2009
}[for experiments \eg][]{guntonetal1983}.

@@

First of all the sampling relation. We have a population of $N$ units --
neurons -- $\yA$ of which have some specific property -- being active in a
specific time bin -- and we sample in an unknown way $n$ of the $N$ units.
The probability that $\ya$ of the $n$ sampled neurons are active is then
given by the hypergeometric distribution
\begin{equation}
  \label{eq:conditional_hypergeometric}
  \yG_{\ya\yA} \defd \pf(\ya \|\yA, n,N)=
  \binom{\yA}{\ya}\binom{N-\yA}{n-\ya}\binom{N}{n}^{-1}
\equiv  \binom{n}{\ya}\binom{N-n}{\yA-\ya}\binom{N}{\yA}^{-1}
\end{equation}
typical of \enquote{drawing without replacement}
\citep[\eg][\chap~3]{jaynes1994_r2003}. %[\sect~4.8.3]{ross1976_r2010}[\sect~II.6]{feller1950_r1968}[\sects~2.1, 3.2]{jeffreys1939_r1983}.
In the following we leave $n$, $N$ implicit in the conditional. If we are
uncertain about the number $\yA$, with belief $P(\yA)$, then by the theorem
of total probability our belief about $\ya$ is
\begin{equation}
  \label{eq:belief_sample_unknown_A}
  p(\ya)= \sum_{\yA} \yG_{\ya\yA} \;P(\yA).
\end{equation}

From the definition of normalized factorial
moment~\eqref{eq:def_factorial_moment_a}, the expression for the
hypergeometric distribution~\eqref{eq:conditional_hypergeometric}, and the
relation~\eqref{eq:belief_sample_unknown_A} between our beliefs about $\ya$
and $\yA$, and using some combinatorial juggling
\citep%[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}
[appendix~A and references therein]{portamanaetal2015}, one can prove the
equality~\eqref{eq:equal_m_factorialmoments} between the factorial moments
of sample and larger population.

@@

For the construction of a maximum-entropy distribution from generic
expectation constraints see Jaynes
\citey{jaynes1963}[\chap~11]{jaynes1994_r2003}, and Hobson \etal\
\citey{hobsonetal1973} and Sivia \citey[\sect~5.2.2]{sivia1996_r2006} for
the minimum-relative-entropy construction. For the solution of the
extremization problem using Lagrangians and Lagrange multipliers see Mead
\etal\ \citey{meadetal1984} and the extensive texts by Fang \etal\
\citey{fangetal1997} and Boyd \etal\ \citey{boydetal2004_r2009}. For a
geometric understanding of the extremization and of the relation between
expectations and multipliers see Porta Mana \citey{portamana2017b}.

@@

The first assumption, of infinite exchangeability
\citep[\chap~4]{bernardoetal1994_r2000}{dawid2013%cifarellietal1982
} of the probability distribution $\pf[(\yA_{1},\dotsc, \yA_{T}) \| \yI]$
for the sequence of activities, implies the following mixture form by
de~Finetti's theorem \citep{definetti1930,hewittetal1955}:
\begin{equation}
  \label{eq:mixture_from_definetti}
%  \exp[-T\;\sh(\yF;\ynu)] \quad
%  \pxxx[(\yA_{1},\dotsc,\yA_{T})] \propto
  \pf[(\yA_{1},\dotsc, \yA_{T}) \| \yI]
  = \int\!\!\!\di\ynu\;\pf(\ynu \|\yI)\,\prod_{\yA=0}^{N}{\nu_{\yA}}^{T\,\yFF_{\yA}},
\end{equation}
where $\ynu\defd(\nu_{\yA})$ is the long-run frequency distribution of
activities in an imaginary continuation or repetition of the recording, and
$\pf(\ynu \|\yI)\;\di\ynu$ our belief distribution about it. The
integration is over an $N$-dimensional simplex.
